CrypTool Script \\ Mathematics and Cryptography

### 2003-04-01 script/en/chapters/aboutcryptool.tex
more fixes

You will find information about the principles of symmetrical and
asymmetrical {\bf encryption}. A large section of the script is dedicated
to the fascinating topic of {\bf prime numbers}. Using numerous examples,
the {\bf elementary number theory} and {\bf modular arithmetic} are
introduced and applied in an exemplary manner to the {\bf RSA procedure}.
After this, one receives insight into the mathematical ideas behind
{\bf modern cryptography}.

@@

Whereas the program CrypTool\index{CrypTool} teaches you how to use
cryptography in practice, the script provides those interested in the
subject with a deeper understanding of the mathematical algorithms
used -- trying to do it didactically as well as possible.

>>>>

You will get information about the principles of symmetrical and
asymmetrical {\bf encryption}. A large section of the script is dedicated
to the fascinating topic of {\bf prime numbers}. Using numerous examples,
the {\bf elementary number theory} and {\bf modular arithmetic} are
introduced and applied in an exemplary manner to the {\bf RSA procedure}.
After this, one receives insight into the mathematical ideas behind
{\bf modern cryptography}.

@@

Whereas the \textit{program} CrypTool\index{CrypTool} teaches you how to use
cryptography in practice, the \textit{script} provides those interested in the
subject with a deeper understanding of the mathematical algorithms
used -- trying to do it didactically as well as possible.

### 2003-04-01 script/en/chapters/cryptomethods.tex
more fixes

For all other procedures there is a (theoretical) possibility of breaking them.
If the procedures are good, however, the time taken to break them is so long
that it is practically impossible to do and these procedures can therefore be
considered (practically) secure.\par \vskip + 3pt

The book of Bruce Schneier \cite{Schneier1996} offers a very good overview about the different algorithms. We basically distinguish between symmetric and asymmetric encryption procedures.

@@

For {\em symmetric} encryption \index{Encryption!symmetric} the sender and
recipient must be in possession of a common (secret) key which they have
exchanged before actually starting to communicate. The sender uses this
key to encrypt the message and the recipient uses it to decrypt it.
\par \vskip + 3pt

@@

The current status on brute-force attacks on symmetric encryption algorithms can be explained with the block cipher RC5.

Brute-force (exhaustive search, trial-and-error) means to completely examine all keys of the key space: so no special analysis methods have to be used. Instead, the cipher text is decrypted with all possible keys and for each resulting text it is checked, whether this is a meaningful clear text. A key length of 64 bit means at most $2^{64}$ = 18,446,744,073,709,551,616 or about 18 trillion (GB) / 18 quintillion (US)  keys to check\index{CrypTool}%
\footnote{%
    With CrypTool\index{CrypTool} v1.3 you also can try brute-force attacks
    of modern symmetric algorithms 
    (using the menu {\bf Analysis \textbackslash{} Ciphertext-Only}): 
    to achieve a result in a appropriate time with a single PC you should 
    mark not more than 20 bit of the key as unknown.
}.

Companies like RSA Security provide so-called cipher challenges in order to quantify the security offered by well-known symmetric ciphers as DES, Triple-DES or RC5\footnote{\href{http://www.rsasecurity.com/rsalabs/challenges/secretkey/index.html}{\tt http://www.rsasecurity.com/rsalabs/challenges/secretkey/index.html}}. They offer prizes for those who manage to decipher cipher texts, encrypted with different algorithms and different key lengths and unveil the symmetric key (under controlled conditions). So theoretical estimates can be confirmed.

@@

If Alice\index{Alice}%
\footnote{%
      In order to describe cryptographic protocols the participants
      are often named Alice, Bob, \dots (see \cite[S. 23]{Schneier1996}). 
      Alice and Bob perform all 
      2-person-protocols. Alice will initiate all protocols and 
      Bob answers. The attackers are named Eve (eavesdropper) and
      Mallory (malicious active attacker).
  } wants to communicate with Bob, then she finds Bob's public key 
in the directory and uses it to encrypt her message to him. She then sends
this cipher text to Bob, who is then able to decrypt it again using his 
secret key. As only Bob knows his secret key, only he can decrypt 
messages addressed to him.
Even Alice who sends the message cannot restore plaintext from the (encrypted)
message she has sent. Of course, you must first ensure that the public key
cannot be used to derive the private key.\par \vskip + 3pt

@@

In this case the data is encrypted using symmetric procedures: the key is a
session key\index{Session key} generated by the sender
randomly\footnote{%
An important part of cryptographically secure techniques is to generate 
random numbers. Within CrypTool\index{CrypTool} v1.3 you can check out
different random number generators using the menu {\bf Indiv. Procedures 
\textbackslash{} Generate Random Numbers}. 
Using the menu{\bf Analysis \textbackslash{} Random Tests} you can apply
different test methods for random data to binary documents. \\
Up to now CrypTool has concentrated on cryptographically strong 
pseudo random number generators. Only the integrated Secude generator 
involves a "pure" random source. 
}\index{Random}
that is only used for this message.
This session key is then encrypted using the asymmetric procedure and
transmitted to the recipient together with the message. Recipients can determine
the session key using their secret keys and then use the session key to encrypt
the message. In this way, we can benefit from the easy key management
\index{Key management} of asymmetric procedures and encrypt large quantities of data
quickly and efficiently using symmetric procedures.

@@

\bibitem[Schmeh2003]{Schmeh2003}  \index{Schmeh 2003}
        Klaus Schmeh, \\
        {\em Cryptography and Public Key Infrastructures on the Internet},\\ 
        John Wiley \& Sons Ltd., Chichester 2003. \\
        A considerable, up-to-date, good reading book, which also 
        considers practical problems like standardisation or
        real existing software.

\bibitem[Schneier1996]{Schneier1996} \index{Schneier 1996} 
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C}, \\
    Wiley 1994, 2nd edition 1996.

>>>>

For all other procedures there is a (theoretical) possibility of breaking them.
If the procedures are good, however, the time taken to break them is so long
that it is practically impossible to do so, and these procedures can therefore be
considered (practically) secure.\par \vskip + 3pt

The book of Bruce Schneier \cite{Schneier1996cm} offers a very good overview about the different algorithms. We basically distinguish between symmetric and asymmetric encryption procedures.

@@

For {\em symmetric} encryption \index{Encryption!symmetric} sender and
recipient must be in possession of a common (secret) key which they have
exchanged before actually starting to communicate. The sender uses this
key to encrypt the message and the recipient uses it to decrypt it.
\par \vskip + 3pt

@@

The current status of brute-force attacks on symmetric encryption algorithms can be explained with the block cipher RC5.

Brute-force (exhaustive search, trial-and-error) means to completely examine all keys of the key space: so no special analysis methods have to be used. Instead, the cipher text is decrypted with all possible keys and for each resulting text it is checked, whether this is a meaningful clear text. A key length of 64 bit means at most $2^{64}$ = 18,446,744,073,709,551,616 or about 18 trillion (GB) / 18 quintillion (US)  keys to check\index{CrypTool}%
\footnote{%
    With CrypTool\index{CrypTool} v1.3 you also can try brute-force attacks
    of modern symmetric algorithms 
    (using the menu {\bf Analysis \textbackslash{} Ciphertext-Only}): 
    to achieve a result in an appropriate time with a single PC you should 
    mark not more than 20 bit of the key as unknown.
}.

Companies like RSA Security provide so-called cipher challenges in order to quantify the security offered by well-known symmetric ciphers as DES, Triple-DES or RC5\footnote{\href{http://www.rsasecurity.com/rsalabs/challenges/secretkey/index.html}{\tt http://www.rsasecurity.com/rsalabs/challenges/secretkey/index.html}}. They offer prizes for those who manage to decipher cipher texts, encrypted with different algorithms and different key lengths, and to unveil the symmetric key (under controlled conditions). So theoretical estimates can be confirmed.

@@

If Alice\index{Alice}%
\footnote{%
      In order to describe cryptographic protocols participants
      are often named Alice, Bob, \dots (see \cite[p. 23]{Schneier1996cm}). 
      Alice and Bob perform all 
      2-person-protocols. Alice will initiate all protocols and 
      Bob answers. The attackers are named Eve (eavesdropper) and
      Mallory (malicious active attacker).
  } wants to communicate with Bob, then she finds Bob's public key 
in the directory and uses it to encrypt her message to him. She then sends
this cipher text to Bob, who is then able to decrypt it again using his 
secret key. As only Bob knows his secret key, only he can decrypt 
messages addressed to him.
Even Alice who sends the message cannot restore plaintext from the (encrypted)
message she has sent. Of course, you must first ensure that the public key
cannot be used to derive the private key.\par \vskip + 3pt

@@

In this case the data is encrypted using symmetric procedures: the key is a
session key\index{Session key} generated by the sender
randomly\footnote{%
An important part of cryptographically secure techniques is to generate 
random numbers. Within CrypTool\index{CrypTool} v1.3 you can check out
different random number generators using the menu {\bf Indiv. Procedures 
\textbackslash{} Generate Random Numbers}. 
Using the menu {\bf Analysis \textbackslash{} Random Tests} you can apply
different test methods for random data to binary documents. \\
Up to now CrypTool has concentrated on cryptographically strong 
pseudo random number generators. Only the integrated Secude generator 
involves a "pure" random source. 
}\index{Random}
that is only used for this message.
This session key is then encrypted using the asymmetric procedure and
transmitted to the recipient together with the message. Recipients can determine
the session key using their secret keys and then use the session key to encrypt
the message. In this way, we can benefit from the easy key management
\index{Key management} of asymmetric procedures and encrypt large quantities of data
quickly and efficiently using symmetric procedures.

@@

\bibitem[Schmeh2003]{Schmeh2003}  \index{Schmeh 2003}
        Klaus Schmeh, \\
        {\em Cryptography and Public Key Infrastructures on the Internet},\\ 
        John Wiley \& Sons Ltd., Chichester 2003. \\
        A considerable, up-to-date, easy to read book, which also 
        considers practical problems like standardisation or
        real existing software.

\bibitem[Schneier1996]{Schneier1996cm} \index{Schneier 1996} 
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C}, \\
    Wiley 1994, 2nd edition 1996.

### 2003-04-01 script/en/chapters/digitalsignatures.tex
be updates
more fixes

\newpage
\section{Digital Signatures}
\index{Signatures!digital}
(Esslinger B. / Filipovics B. / Schneider J. / Koy H., June 2002,
Update: Feb. 2003)

@@

The document is sent to the recipient together with the signature. The recipient
can then use the sender's public key, the document and the signature to
establish whether or not the signature is correct. In practice, however, the
procedure we have just described has a decisive disadvantage. The signature is
approximately as long as the document itself. To prevent an unnecessary increase
in data traffic, and also for reasons of performance\index{Performance}, we use 
a cryptographic hash function\footnote{%
%xxxxxxxxxx Menüeinträgsbezeichnungen ok?
Hash functions\index{Hash function} are implemented within CrypTool\index{CrypTool} v1.3.04 at
several places. \\
Using menu {\bf Indiv. Procedures \textbackslash{} Hash} you have the possibilities
\begin{itemize}\addtolength{\itemsep}{-1.0\baselineskip} % um den Abstand zu verringern bebe
\item to apply one of 6 hash functions to the content of the current window, \\
\item to calculate the hash value of a file, \\
\item to test, how changes to a text change the according hah value and \\
\item to perform a simulation, how digital signatures could be attacked 
      by a targeted search for hash value collisions.
\end{itemize}
}
.

@@

While DSA was specifically designed, such that it can be exported from
countries regulating export of encryption soft and hardware (like the U.S.\ 
at the time when it was specified), it has been noted
\cite[p.~490]{5Schneier1996}, that the operations involved in DSA can be
used to emulate RSA and ElGamal encryption.

@@

\bibitem[Schneier1996]{5Schneier1996} \index{Schneier 1996} 
    Bruce Schneier, \\
    Applied Cryptography, Protocols, Algorithms, and Source Code in C, \\
    Wiley, 2nd edition, 1996.

>>>>

\newpage
\section{Digital Signatures}
\index{Signatures!digital}
(Schneider J. / Esslinger B. / Filipovics B. / Koy H., June 2002,
Update: Feb. 2003)

@@

The document is sent to the recipient together with the signature. The recipient
can then use the sender's public key, the document and the signature to
establish whether or not the signature is correct. In practice, however, the
procedure we have just described has a decisive disadvantage. The signature is
approximately as long as the document itself. To prevent an unnecessary increase
in data traffic, and also for reasons of performance\index{Performance}, we use 
a cryptographic hash function\footnote{%
%xxxxxxxxxx Menüeinträgsbezeichnungen ok?
Hash functions\index{Hash function} are implemented within CrypTool\index{CrypTool} v1.3.04 at
several places. \\
Using menu {\bf Indiv. Procedures \textbackslash{} Hash} you have the possibilities
% hier die items nicht einrücken!
\begin{list}{\textbullet}{\leftmargin10pt\addtolength{\itemsep}{-1.0\baselineskip}}
%\begin{itemize}\addtolength{\itemsep}{-1.0\baselineskip} % um den Abstand zu verringern bebe
\item to apply one of 6 hash functions to the content of the current window, \\
\item to calculate the hash value of a file, \\
\item to test, how changes to a text change the according hash value and \\
\item to perform a simulation, how digital signatures could be attacked 
      by a targeted search for hash value collisions.
\end{list}
}.

@@

While DSA was specifically designed, such that it can be exported from
countries regulating export of encryption soft and hardware (like the U.S.\ 
at the time when it was specified), it has been noted
\cite[p.~490]{Schneier1996ds}, that the operations involved in DSA can be
used to emulate RSA and ElGamal encryption.

@@

\bibitem[Schneier1996]{Schneier1996ds} \index{Schneier 1996} 
    Bruce Schneier, \\
    Applied Cryptography, Protocols, Algorithms, and Source Code in C, \\
    Wiley, 2nd edition, 1996.

### 2003-04-01 script/en/chapters/introduction.tex
more fixes

In order to also enable developers/authors to work independently together 
the topics have been split up and for each topic an extra chapter was 
written readable on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes, where you
can find the according 
functions within the CrypTool\index{CrypTool} program. Naturally 
there are much more interesting topics in mathematics and cryptography 
which could be more expanded -- therefore this is only one of many
ways to do it.

Rapid spread of the Internet has also lead to an intensified research 
in the technologies involved, especially within the area of cryptography 
where a good deal of new knowledge came up.

This edition of the script mainly updates the summaries of the following 
topical research areas:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers (generalized Mersenne 
        and Fermat primes) \\ (chap. 2.6), 
  \item progress in number theory ("Primes in P") (chap. 3.11.3), 
  \item the factorisation of big numbers (TWIRL) (chap. 3.11.3),
  \item progress in cryptanalysis of the AES standard (chap. 1.1) and
  \item progress in brute-force attacks against symmetric algorithms (chap. 1.1).
\end{itemize}

@@

The first time the script was delivered with CrypTool\index{CrypTool} 
was in version 1.2.01. Since then it has been extended and updated with each 
new version of CrypTool (1.2.02, 1.3.00, 1.3.02, 1.3.03 and now 1.3.04).

@@

% --------------------------------------------------------------------------
\newpage
\section*{Introduction -- How do the Script and CrypTool Play together?}  \addcontentsline{toc}{section}{Introduction -- How do the Script and CrypTool Play together?}

@@

CrypTool\index{CrypTool} was developed during the end-user awareness program
in order to increase employee awareness of IT security and provide them with
a deeper understanding of the term security.\par \vskip + 3pt

A further aim has been to enable users to understand the cryptographic
procedures implemented in the Deutsche Bank. In this way, using CrypTool
as a reliable reference implementation of the various encryption procedures
(because of using the industry-proven Secude Library\index{Secude}),
you can test the encryption implemented in other programs. \par \vskip + 3pt

Since then CrypTool\index{CrypTool} has often been used for 
education in companies and teaching at school and universities and
moreover several universities are helping to further develop the project.
\par \vskip + 15pt

>>>>

In order to also enable developers/authors to work together independently 
the topics have been split up and for each topic an extra chapter was 
written readable on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes describing where you
can find the according 
functions within the CrypTool\index{CrypTool} program. Naturally 
there are much more interesting topics in mathematics and cryptography 
which could be more expanded -- therefore this is only one of many
ways to do it.

\vspace{12pt}
Rapid spread of the Internet has also lead to an intensified research 
in the technologies involved, especially within the area of cryptography 
where a good deal of new knowledge came up.

This edition of the script mainly updates the summaries of the following 
topical research areas:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers (generalized Mersenne 
        and Fermat primes) \\ (chap. 2.6), 
  \item progress in number theory ("Primes in P") (chap. 3.11.3), 
  \item the factorisation of big numbers (TWIRL) (chap. 3.11.3),
  \item progress in cryptanalysis of the AES standard (chap. 1.1.1) and
  \item progress in brute-force attacks against symmetric algorithms (chap. 1.1.2).
\end{itemize}

@@

\vspace{12pt}
The first time the script was delivered with CrypTool\index{CrypTool} 
was in version 1.2.01. Since then it has been extended and updated with each 
new version of CrypTool (1.2.02, 1.3.00, 1.3.02, 1.3.03 and now 1.3.04).

@@

% --------------------------------------------------------------------------
\newpage
\section*{Introduction -- How do the Script and the Program Play together?}  \addcontentsline{toc}{section}{Introduction -- How do the Script and the Program Play together?}

@@

CrypTool\index{CrypTool} was developed during the end-user awareness program
at Deutsche Bank in order to increase employee awareness of IT security and provide them with
a deeper understanding of the term security.\par \vskip + 3pt

A further aim has been to enable users to understand the cryptographic
procedures. In this way, using CrypTool
as a reliable reference implementation of the various encryption procedures
(because of using the industry-proven Secude Library\index{Secude}),
you can test the encryption implemented in other programs. \par \vskip + 3pt

CrypTool\index{CrypTool} has often been used for 
education in companies and teaching at school and universities, and
moreover several universities are helping to further develop the project.
\par \vskip + 15pt

### 2003-04-01 script/en/chapters/moderncryptography.tex
be updates
more be updates
more fixes

\newpage
%--------------------------------------------------------------------
\subsection{Knapsack problem as a basis for public key procedures}%
\index{Cryptography!Public Key}
%--------------------------------------------------------------------
%--------------------------------------------------------------------
\subsubsection{Knapsack problem} \index{Knapsack}
%--------------------------------------------------------------------

@@

\subsection{Decomposition into prime factors as a basis for public key 
procedures}

@@

Let $ p $ and $ q $ be two different prime numbers and $ N = pq$. Let 
$ e $ be any prime number relative to $ \phi (N) $ \index{Prime number!relative}, i.e. $ \ggt (e,\phi (N))=1. $\index{Gcd}
Using the Euclidean algorithm, 
we calculate the natural number  $ d < \phi (N), $ such that
\[ ed \equiv 1 \mod \phi (N). \]
whereby $ \phi $ is the {\bf Euler phi Function}.

>>>>

\newpage
%--------------------------------------------------------------------
\subsection{Knapsack problem as a basis for public key procedures}%
\index{Cryptography!public key}
%--------------------------------------------------------------------
%--------------------------------------------------------------------
\subsubsection{Knapsack problem} \index{Knapsack}
%--------------------------------------------------------------------

@@

\subsection{Decomposition into prime factors as a basis for public key 
procedures}\index{Prime factor!decomposition}

@@

Let $ p $ and $ q $ be two different prime numbers and $ N = pq$. Let 
$ e $ be any prime number relative to $ \phi (N) $ \index{Prime number!relative prime}, i.e. $ \ggt (e,\phi (N))=1. $\index{Gcd}
Using the Euclidean algorithm, 
we calculate the natural number  $ d < \phi (N), $ such that
\[ ed \equiv 1 \mod \phi (N). \]
whereby $ \phi $ is the {\bf Euler phi Function}.

### 2003-04-01 script/en/chapters/numbertheory.tex
more fixes

% --------------------------------------------------------------------------
\subsubsection{Calculation of the multiplicative Inverse}

@@

{\bf Example:}\par
Find the multiplicative inverse of $1579$ modulo $7351$.\\
According to Euler-Fermat:  $a^{J(n)} = 1 {\rm ~(mod~} n)$ for all $a$ in $\mathbb{Z}_n^*$.
If we divide both sides by $a$, we get: $a^{J(n) - 1} \equiv a^{-1} {\rm ~(mod~} n)$.
For the special case that the modulo is prime, we have $J(n) = p - 1$.
Therefore, the modular inverse is 
$$a^{-1} = a^{J(n) - 1} \equiv a^{(p-1)-1} \equiv a^{p-2} {\rm ~(mod~} p).$$
For our example, this means:
\begin{itemize}
\item[] Since the modulus $7351$ is prime, $p-2 = 7349$. \\
    $1579^{-1} \equiv 1579^{7349} {\rm ~(mod~} p).$
\end{itemize}
By cleverly breaking down the exponent, we can calculate this power relatively easily (see Section~\ref{hohpot}
\hyperlink{hohpot}{Fast calculation of high powers)}:
\begin{itemize}
\item[] $7349 = 4096 + 2048 + 1024 + 128 + 32 + 16 + 4 + 1$
\item[] $1579^{-1} \equiv 4716 {\rm ~(mod~} 7351).$
\end{itemize}

@@

This record, established on January 18, 2002 by researchers of the German
university of Bonn\footnote{%
\href{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}
{\texttt{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}}}, got much less attention within the press than the
solution of the RSA challenge, established on August 22, 1999, when
researchers from the Netherlands factorized the 155-digit number into its
both 78-digit primes (see Chapter \ref{chptSecurityParam}). 
One reason is that 512 bit RSA-155 was a kind of {\em magic} border.

@@

Alice and Bob\footnote{Bob and Alice are the default names used for the two authorised participants in a protocol 
(see \cite[p. 23]{Schneier1996}).
} use a one way function to obtain a key $S$, the session key, for subsequent correspondence. This is then a secret 
that is only known to the two of them. Alice selects a random number $a$ and keeps it secret. She applies a one way 
function to $a$ to calculate the number $A = g^a$ and sends it to Bob. He does the same, by selecting a secret random 
number $b$, calculating $B = g^b$ and sending it to Alice. The number $g$ is random and can be publicly known. 
Alice applies the one way function together with her secret number $a$ to $B$, while Bob does the same with his 
secret number $b$ and the received number $A$.

@@

{\bf Example using numbers:} 
\begin{itemize}
\item[{\bf 1.}] Alice and Bob select $g = 11, p = 347$.
\item[{\bf 2.}] Alice selects $a = 240$, Bob selects $b = 39$ and they keep $a$ and $b$ secret.
\item[{\bf 3.}] Alice calculates $A \equiv g^a \equiv 11^{240}  \equiv 49 {\rm ~(mod~} 347).$\\
                Bob calculates $B \equiv g^b \equiv 11^{39} \equiv 285 {\rm ~(mod~} 347).$
\item[{\bf 4.}] Alice sends Bob:   $A \equiv 49$,\\
                Bob sends Alice: $B \equiv 285$.
\item[{\bf 5.}] Alice calculates $B^a \equiv 285^{240} \equiv 268 {\rm ~(mod~} 347),$\\
                Bob calculates $A^b \equiv   49^{39} \equiv 268 {\rm ~(mod~} 347).$
\end{itemize}
Alice and Bob can now communicate securely using their shared session key. Even if spies were to intercept everything 
transferred via the connection:
$g = 11, p = 347, A = 49$ and $B = 285$, they would not be able to calculate the secret key.

@@

\bibitem[GISA2002]{GISA2002} \index{GISA 2002}
    GISA (German Information Security Agency) \\
    {\em Recommendation for key length selection}, Bonn, Sep. 2002, \\
    \href{http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf}
    {\tt http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf} \\
    A statement on these recommendations: \\
    \href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
    {\texttt{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}}

@@

\bibitem[Schneier1996]{Schneier1996} \index{Schneier 1996} 
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C}, \\
    Wiley and Sons, 2nd edition 1996.

>>>>

% --------------------------------------------------------------------------
\subsubsection{Calculation of the multiplicative inverse}

@@

{\bf Example:}\par
Find the multiplicative inverse of $1579$ modulo $7351$.\\
According to Euler-Fermat:  $a^{J(n)} = 1 {\rm ~(mod~} n)$ for all $a$ in $\mathbb{Z}_n^*$.
If we divide both sides by $a$, we get: $a^{J(n) - 1} \equiv a^{-1} {\rm ~(mod~} n)$.
For the special case that the modulo is prime, we have $J(n) = p - 1$.
Therefore, the modular inverse is 
$$a^{-1} = a^{J(n) - 1} \equiv a^{(p-1)-1} \equiv a^{p-2} {\rm ~(mod~} p).$$
For our example, this means:
\begin{itemize}
\item[] Since the modulus $7351$ is prime, $p-2 = 7349$. \\
    $1579^{-1} \equiv 1579^{7349} {\rm ~(mod~} p).$
\end{itemize}
By cleverly breaking down the exponent, we can calculate this power relatively easily (see Section~\ref{hohpot}
\hyperlink{hohpot}{Fast calculation of high powers}):
\begin{itemize}
\item[] $7349 = 4096 + 2048 + 1024 + 128 + 32 + 16 + 4 + 1$
\item[] $1579^{-1} \equiv 4716 {\rm ~(mod~} 7351).$
\end{itemize}

@@

This record, established on January 18, 2002 by researchers of the German
University of Bonn\footnote{%
\href{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}
{\texttt{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}}}, got much less attention within the press than the
solution of the RSA challenge\footnote{\href{http://www.rsasecurity.com/rsalabs/challenges/factoring/numbers.html}
         {\texttt{http://www.rsasecurity.com/rsalabs/challenges/factoring/numbers.html}}}, established on August 22, 1999, when
researchers from the Netherlands factorized the 155-digit number into its
both 78-digit primes (see Chapter \ref{chptSecurityParam}). 
One reason is that 512 bit RSA-155 was a kind of {\em magic} border.

@@

Alice and Bob\footnote{Bob and Alice are the default names used for the two authorised participants in a protocol 
(see \cite[p. 23]{Schneier1996nt}).
} use a one way function to obtain a key $S$, the session key, for subsequent correspondence. This is then a secret 
that is only known to the two of them. Alice selects a random number $a$ and keeps it secret. She applies a one way 
function to $a$ to calculate the number $A = g^a$ and sends it to Bob. He does the same, by selecting a secret random 
number $b$, calculating $B = g^b$ and sending it to Alice. The number $g$ is random and can be publicly known. 
Alice applies the one way function together with her secret number $a$ to $B$, while Bob does the same with his 
secret number $b$ and the received number $A$.

@@

\pagebreak
{\bf Example using numbers:} 
\begin{itemize}
\item[{\bf 1.}] Alice and Bob select $g = 11, p = 347$.
\item[{\bf 2.}] Alice selects $a = 240$, Bob selects $b = 39$ and they keep $a$ and $b$ secret.
\item[{\bf 3.}] Alice calculates $A \equiv g^a \equiv 11^{240}  \equiv 49 {\rm ~(mod~} 347).$\\
                Bob calculates $B \equiv g^b \equiv 11^{39} \equiv 285 {\rm ~(mod~} 347).$
\item[{\bf 4.}] Alice sends Bob:   $A \equiv 49$,\\
                Bob sends Alice: $B \equiv 285$.
\item[{\bf 5.}] Alice calculates $B^a \equiv 285^{240} \equiv 268 {\rm ~(mod~} 347),$\\
                Bob calculates $A^b \equiv   49^{39} \equiv 268 {\rm ~(mod~} 347).$
\end{itemize}
Alice and Bob can now communicate securely using their shared session key. Even if spies were to intercept everything 
transferred via the connection:
$g = 11, p = 347, A = 49$ and $B = 285$, they would not be able to calculate the secret key.

@@

\bibitem[GISA2002]{GISA2002} \index{GISA 2002}
    GISA (German Information Security Agency) \\
    {\em Recommendation for key length selection}, Bonn, Sep. 2002, \\
    \href{http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf}
    {\tt http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf} \\
    A statement on these recommendations: \\
    \href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
    {\texttt{http://www.secorvo.de/publikat/\\
    \vspace*{1cm}stellungnahme-algorithmenempfehlung-020307.pdf}}

@@

\bibitem[Schneier1996]{Schneier1996nt} \index{Schneier 1996} 
    Bruce Schneier, \\
    {\em Applied Cryptography, Protocols, Algorithms, and Source Code in C}, \\
    Wiley and Sons, 2nd edition 1996.
